Estimating Predictive Rate-Distortion via Neural Variational Inference

### 2019-02-02 paper.tex
Updates from ShareLaTeX

\newcommand{\soft}[1]{}
\newcommand{\nopreview}[1]{}
\newcommand\comment[1]{{\color{red}#1}}
\newcommand\mhahn[1]{{\color{red}(#1)}}

@@

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}

@@

%\emnlpfinalcopy
%\def\emnlppaperid{496}

\title{Estimating Predictive Rate-Distortion via Neural Variational Inference} %Crosslinguistic Word Orders Trade Off Predictability and Memory Load}%  Word Order Minimizes  Across Languages
\author{Michael Hahn}
\date{2018}

@@

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. % and \emph{decoder} $\psi$ 
The encoder transforms observation sequences $x_{-M\dots -1}$ into a distribution over $Z \in \mathbb{R}^N$: 
\begin{equation}
	P(Z|X_{-M\dots M}) := \phi(Z|X_{-M\dots-1}) := \phi(X_{-M\dots-1})(Z)
\end{equation}
From this, one can obtain the conditional distribution over future observations via Bayes' rule, using the Markov condition:
\begin{equation}\label{eq:cond-fut}
	P(X_{0\dots M-1}|Z) = \frac{\E_{X_{-M \dots M-1}} \phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})}
\end{equation}
However, if $\phi$ is given by a powerful function approximating family such as recurrent neural networks, neither of the two expectations is tractable,  as they require summation over exponentialy many sequences.

@@

If we fix a stochastic process $(X_t)_t$ and an encoder $\phi$, then -- for \emph{any} choice of these functions $\psi$ and $q$ -- the following two bounds hold.
First, the loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact conditional (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(X_{0\dots M-1} | z)\right] \geq \operatorname{H}[X_{1\dots M-1}|Z]
\end{equation}
Furthermore, equality is attained if and only if $\psi(X_{0\dots M-1}|Z) = P(X_{0\dots M-1}|Z)$.
%It is in this sense that $\psi$ `approximates' the conditional $P(X_{0\dots M-1}|Z)$.

\begin{proof}
%\begin{equation}\label{ineq1}
\begin{align*}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(X_{0\dots M-1} | z)\right] & \geq -	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P(X_{0\dots M-1} | z)\right]\\
	&= \operatorname{H}[X_{1\dots M-1}|Z]
\end{align*}
%\end{equation}
\end{proof}

Second, the KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ D_{KL}(P(Z|X) || q(Z))\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = I[X_{-M\dots 0},Z] 
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.

%These inequalities are tight if $\psi$ is the Bayes optimal decoder ($\psi_z(y) = \sum_{x} \phi(z|x) P(x, y)$), and $q$ is the marginal of $Z$.
% N(\mu(x), \sigma^2(x))
%Note that, for finite $M$, the predictive rate-distortion objective is equivalent to
Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \beta \cdot D_{KL}(\phi(X_{-M\dots-1}) || q)\right]
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
%As this optimization ranges over all distributions $\phi, \psi, q$ of the appropriate domains, t
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and -- if the family of approximators is sufficiently rich -- can come close to turning these into equalities.
%Optimizing them for the objective (\ref{eq:bound}) will result in functions that approximate the optimal decoder and the marginal.
That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal.

@@

For the deocder $\psi$, we use a recurrent LSTM network to compute a sequence of vector representations $g_t$ from the code $z$ and the prefix $X_{1\dots t-1}$ for $t = 1, \dots T$, and set
\begin{equation}
	\psi(x_t = s_i|x_{1...t-1}, z) \propto \exp((W_o g_t)_i)
\end{equation}
where $s_i$ is an element of the state space $S$, indexed by $i=1, ..., |S|$, and $W_o \in \mathbb{R}^{k \times |S|}$ is a parameter.

@@

\paragraph{Optimisation}
We optimize the parameters of the neural networks expressing $\phi, \psi, q$ for (\ref{eq:bound}) using Backpropagation and Adam \citep{kingma-adam:-2014}, a standard and widely used method for optimizing neural networks.
During optimization, we approximate the gradient by taking a single sample from $z$ per sample trajectory and use the reparametrized gradient estimator introduced by \citet{kingma-auto-encoding-2014}.
%\begin{equation}\label{eq:bound-mc}
%	\partial_{\phi, \psi, q}	\log P(X_{0\dots T} | z_{-M...0}) + \beta \cdot \log \frac{N(\mu, \sigma^2)(z)}{q(z_{-M...0})}
%\end{equation}
%for a given training example $x_{-M...M}$.
This results in an unbiased estimator of the gradient of (\ref{eq:bound}) w.r.t. the parameters of $\phi, \psi, q$.

We follow standard practice in Machine Learning by splitting the data into a training and a held-out set, using the training set for optimising the parameters.
After every go through the training set, the objective is evaluated on the held-out set; optimisation terminates once the value on the held-out set does not improve any more.

\paragraph{Evaluating}
After optimizing the parameters on a set of observed trajectories, we evaluate it on a held-out set of trajectories to estimate (\ref{eq:bound}).
Given parameters for $\phi, \psi, q$, we evaluate the objective~(\ref{eq:bound}) on the dataset by taking, for each timeseries $X_{-M}...X_M$, a single sample of $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing
\begin{equation}\label{eq:bound-mc}
	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \beta \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
\end{equation}
When evaluating (\ref{eq:bound}) on unseen data this way, the resulting value is guaranteed to be an \emph{upper} bound on the true predictive rate-distortion due to (\ref{ineq1}-\ref{ineq2}), up to sampling error introduced by sampling $z$ and the finite size of the held-out dataset.
It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc}) is an \emph{unbiased} estimator of an upper bound.
The estimation error can be estimated using standard methods such as bootstrapping, cross-validation, and taking more samples of $z$ -- unlike overfitting affecting OCE, which \emph{biases} the values obtained.
%The use of a held-out set is motivated by the fact that -- as for OCE -- the codebook may have overfitted to the set used for optimization.

@@

NPRD provably provides \emph{upper bounds} on the predictive rate-distortion curve.
This makes it possible to objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve.

@@

%We optimize $q(z|x)$, marginal $s(z)$, and $r(y|z)$.
\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used in our neural method:
We use held-out data to compute
$$\frac{1}{N} \sum_{i=1}^N D_{KL}(q(\cdot|x_i) || s(\cdot))$$ %\E_{z \sim q(\cdot|x_i)} \log \frac{q(z|x_i)}{s(z)} $$
and
$$\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z)$$
Thanks to (\ref{ineq1}-\ref{ineq2}), this provides upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.
In order to be able to apply OCE on held-out data, which will contain sequences that did not occur during estimation, we add a pseudo-sequence $\omega$ and add pseudo-observations $(\omega, x), (x, \omega), (\omega, \omega)$ for all observed sequences $x_{-M...0}, x_{1...M}$ to the matrix of observed counts that serves as the input to the Blahut-Arimoto algorithm.
These pseudo-observations were assigned pseudo-counts $\gamma$ in this matrix of observed counts; we found that a wide range of values ranging from $0.0001$ to $1.0$ yielded essentially the same results.
When evaluating the codebook on held-out data, novel sequences were mapped to $\omega$.
%Given that all unseen futures were mapped to $\omega$, this method may slightly bias towards underestimating entropy of the

@@

Neural networks have hyperparameters, such as the number of units and the step size used in optimization, that affect the quality of approximation depending on properties of the dataset and the function beig approximated.
Given that NPRD provably provides upper bounds, one can identify the best hyperparameters for a given problem by choosing the combination that leads to the lowest estimated upper bounds.

@@

\begin{tabular}{llll}
	            & Tractable \\
	Dropout rate & 0.0, 0.1, 0.2 \\
	Embedding Dimension & 50, 100 \\
	LSTM Dimension & 32 \\
	LSTM Layers & 1 \\
	Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004 \\
	Input Dropout & 0 \\
	Batch Size & 16, 32, 64 \\
	Flow Length & 1, 2 \\
	Flow Type & DSF, DDSF \\
	Flow Dimension & 32, 64, 128, 512 \\
	Flow Layers & 2 \\
\end{tabular}

@@

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \emph{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of subsequent ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \cite{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
%Their method using the (already known) epsilon machine  works well.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \emph{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of ones with blocks of at most two zeros interspersed.
The $\epsilon$-machine is given in Figure~\ref{fig:rip-machine} (right).

@@

\begin{figure*}
\includegraphics[width=0.45\textwidth]{cryptfig/even-info.pdf}
\includegraphics[width=0.45\textwidth]{cryptfig/rip-info.pdf}

%\includegraphics[width=0.45\textwidth]{cryptfig/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process and the Random Insertion Process. Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Blue: OCE for $M\leq 5$. }\label{fig:even}
\end{figure*}

@@

@@

In Figure~\ref{fig:rip-machine}, we record, for each of the three modes where the distribution of the code $Z$ shifts when a symbol is appended.
For comparison, we show the $\epsilon$-machine computed by \cite{marzen-predictive-2016}.
Comparing the two diagrams shows that NPRD effectively recovers the $\epsilon$-machine. % even though $Z$ is a multivariate Gaussian variable.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{cryptfig/foo_pca_2.png}
\includegraphics[width=0.45\textwidth]{cryptfig/foo_pca_3.png}
	\caption{Applying PCA to 5000 sampled codes $Z$ for the Random Insertion Process, at $\beta = 0.25$ and $\beta = 0.6$. Samples are colored according to the states in the $\epsilon$ machine. Black samples belong to sequences that, at $M=15$, cannot be uniquely attributed to any of the states (ambiguous between A and C).}\label{fig:latent}
\end{figure*}
% colors = [{"A" : "red", "B" : "green", "C" : "blue", "U" : "black"}[x] for x in states]

\begin{figure*}
\includegraphics[width=0.45\textwidth]{cryptfig/foo_pca_3_machine.png}
\begin{tikzpicture}[shorten >=1pt,node distance=4cm,on grid,auto] 
   \node[state, fill=green] (B)   {$B$}; 
   \node[state, fill=red] (A) [below left=of B] {$A$}; 
   \node[state, fill=blue] (C) [below right=of B] {$C$}; 
    \path[->] 
	(A) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (B)
	(A) edge [bend left=20] node [below] {$\frac{1}{2} | 1$} (C)
	(B) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (C)
	(B) edge [bend right=20] node [left] {$\frac{1}{2} | 1$} (C)
	(C) edge [bend left=20] node [left] {$\frac{1}{2} | 1$} (A)
	;
%
%%          edge  node [swap] {1} (B)
%    (C) edge  node  {1} (A)
%          edge [loop above] node {0} ()
%	(A) edge  node [swap] {0} (C) 
%          edge [loop below] node {1} ();
\end{tikzpicture}

@@

We consider the following process, which we'll call \textsc{Copy3}: $X_{-15}, ..., X_{-1}$ are independent uniform draws from $\{1,2,3\}$, and $X_1 = X_{-1}, ..., X_{15} = X_{-15}$.
This process deviates a bit from our usual setup since we defined it only for $t \in \{-15, ..., 15\}$, but it is well-suited to investigating this question:
The number of causal states is $3^{15} \approx 14$ Million.
With exactly the same setup as for the \textsc{Even} and \textsc{RIP} processes, NPRD achieved essentially zero distortion on unseen data, even though the number of training samples (3 Million) was far lower than the number of distinct causal states.
However, we found that, at this setup, NPRD overestimated the rate.
Increasing the number of training samples from 3M to 6M, NPRD recovered codebooks that achieved both almost zero distortion and almost optimal rate, on fresh samples (Figure~\ref{fig:repeat}).
Even then, the number of distinct causal states is more than twice the number of training samples.
These results demonstrate that, by using function approximation, NPRD is capable of extrapolating to unseen causal states, encoding and decoding appropriate codes on the fly.

Note that one could easily design an optimal decoder and encoder for \textsc{Copy3} by hand -- the point of this experiment is to demonstrate that NPRD is capable of inducing such a codebook purely from data, in a general-purpose, off-the-shelf manner.
This contrasts with OCE:
Without optimizations specific to the task at hand, a direct application of OCE would require brute-force storing of all 14 Million distinct pasts and futures.

@@

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{cryptfig/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{cryptfig/repeat3-ee-mem-frontier.pdf}
	\caption{Left: Rate-Distortion for the \textsc{Copy3} process. Green: Curve found by NPRD. Gray: Anaytical curve.}\label{fig:repeat}
\end{figure*}

@@

We consider the problem of estimating rate-distortion for Natural Language.
%
%As (CITE) have noted, evaluating (X) on finite data will overestimate MI with the future.
%To address this, we estimate (X) on held-out data.
%
%We compare our estimator with OCF on (1) POS-level language modeling, (2) word-level language modeling
%
%Challenges for OCF:
%
%- held-out data will have unobserved sequences.
%We add an out-of-vocabulary pseudo-sequence to the table of observed sequences.
%

\subsection{POS-level language modeling}

>>>>

\newif \ifcomment
\commentfalse

\newcommand{\soft}[1]{}
\newcommand{\nopreview}[1]{}
\newcommand\comment[1]{\ifcomment{{\color{red}#1}}\else{}\fi}
\newcommand\mhahn[1]{\ifcomment{{\color{red}(#1)}}\else{}\fi}
\newcommand\rljf[1]{\ifcomment{{\color{blue}(#1)}}\else{}\fi}

@@

\newcounter{theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{question}[theorem]{Question}
\newtheorem{example}[theorem]{Example}
\newtheorem{defin}[theorem]{Definition}

@@

\author{Michael Hahn, Richard Futrell}

\title{Estimating Predictive Rate-Distortion via Neural Variational Inference}
\date{2018}

@@

% Some other stuff to cite/consider in framing
% - the Tim Genewein papers on bounded rationality -- predictive rate distortion is like the action of an boundedly rational predictor 
% - Rubin, Shamir & Tishby (2010). Trading value and information in MDPs. Is predictive rate distortion the same as their model but with KL divergence as the value function?

% Possible venues
% - Entropy, of course. 
% - Also the journals that the previous PRD papers showed up in: Chaos, Journal of Statistical Physics, Physical Review Letters

@@

Let us assume that an \emph{encoder} $\phi$ is given, expressed in a family of function approximators. % and \emph{decoder} $\psi$ 
The encoder transforms observation sequences $x_{-M\dots -1}$ into a distribution over $Z \in \mathbb{R}^N$: \rljf{Slight mismatch of the math and the prose here: the prose makes it sound like $\phi$ is going to be a function from $x_{-M\dots -1}$ to $Z$ but in the math it returns a probability mass/density function. Maybe write as Z = \phi($x_{-M\dots -1}$), P_\phi(Z|$x_{-M\dots -1}$) = ...}
\begin{equation}
	P(Z|X_{-M\dots M}) := \phi(Z|X_{-M\dots-1}) := \phi(X_{-M\dots-1})(Z)
\end{equation}
From this, one can obtain the conditional distribution over future observations via Bayes' rule, using the Markov condition: \rljf{Show some intermediate steps here}
\begin{equation}\label{eq:cond-fut}
	P(X_{0\dots M-1}|Z) = \frac{\E_{X_{-M \dots M-1}} \phi(Z|X_{-M\dots-1})}{\E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})}
\end{equation}
However, if $\phi$ is given by a powerful function approximating family such as recurrent neural networks, neither of the two expectations is tractable,  as they require summation over exponentially many sequences. \rljf{Why does this only happen for powerful function approximators? Isn't it the same number of sequences no matter the approximator? Ie doensn't it depend on $M$, not $\phi$?}

@@

If we fix a stochastic process $(X_t)_t$ and an encoder $\phi$, then---for \emph{any} choice of these functions $\psi$ and $q$---the following two bounds hold.

\begin{proposition}
The loss incurred when predicting the future from $Z$ via $\psi$ upper-bounds the true conditional entropy of the future given $Z$, when predicting using the exact conditional (\ref{eq:cond-fut}):
\begin{equation}\label{ineq1}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(X_{0\dots M-1} | z)\right] \geq \operatorname{H}[X_{1\dots M-1}|Z]
\end{equation}
Furthermore, equality is attained if and only if $\psi(X_{0\dots M-1}|Z) = P(X_{0\dots M-1}|Z)$.
%It is in this sense that $\psi$ `approximates' the conditional $P(X_{0\dots M-1}|Z)$.
\end{proposition}

\begin{proof}
%\begin{equation}\label{ineq1}
\begin{align*}
	-	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P_\psi(X_{0\dots M-1} | z)\right] & \geq -	\mathbb{E}_{X}\mathbb{E}_{Z \sim \phi(X)}\left[\log P(X_{0\dots M-1} | z)\right]\\
	&= \operatorname{H}[X_{1\dots M-1}|Z].
\end{align*}
%\end{equation}
\end{proof}

\begin{proposition} 
The KL Divergence between $P(Z|X)$ and $q(Z)$, averaged over $X$, upper-bounds the mutual information between $Z$ and the past observations:
\begin{equation}\label{ineq2}
\begin{aligned}
	\mathbb{E}_{X}\left[ D_{KL}(P(Z|X) || q(Z))\right] &=   \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P_\phi(Z|X_{-M\dots -1})}{q(Z)}  \\
	& \geq  \mathbb{E}_{X} \mathbb{E}_{Z|X} \log \frac{P(Z|X_{-M\dots -1})}{P(Z)}  \\
	& = I[X_{-M\dots 0},Z].
\end{aligned}
\end{equation}
Equality is attained if and only if $q(Z)$ is equal to the true marginal $P(Z) = \E_{X_{-M \dots -1}} \phi(Z|X_{-M \dots -1})$.
\end{proposition}

%These inequalities are tight if $\psi$ is the Bayes optimal decoder ($\psi_z(y) = \sum_{x} \phi(z|x) P(x, y)$), and $q$ is the marginal of $Z$.
% N(\mu(x), \sigma^2(x))
%Note that, for finite $M$, the predictive rate-distortion objective is equivalent to
Therefore, the predictive rate-distortion objective is equivalent to
\begin{equation}\label{eq:bound}
	\min_{\phi, \psi, q}	\left[-	\mathbb{E}_{z \sim \phi(X_{-M\dots-1})}\left[\log P(X_{0\dots M} | z)\right] + \beta \cdot D_{KL}(\phi(X_{-M\dots-1}) || q)\right],
\end{equation}
where $\phi, \psi, q$ range over all triples of the appropriate types described above.
%As this optimization ranges over all distributions $\phi, \psi, q$ of the appropriate domains, t
To carry out this optimization, we will restrict to a powerful family of parametric families of function approximators, within which we will optimize the objective with gradient descent.
The values obtained for the resulting $\phi, \psi, q$ satisfy the bounds in (\ref{ineq1}-\ref{ineq2}), and---if the family of approximators is sufficiently rich---can come close to turning these into equalities.
%Optimizing them for the objective (\ref{eq:bound}) will result in functions that approximate the optimal decoder and the marginal.
That is, we introduce $\psi$ and $q$ as \emph{variational approximations} \citep{blei-variational-2016} to the true Bayes optimal decoder and marginal. \rljf{Expand last sentence into a paragraph}

@@

For the decoder $\psi$, we use a recurrent LSTM network to compute a sequence of vector representations $g_t$ from the code $z$ and the prefix $X_{1\dots t-1}$ for $t = 1, \dots T$, and set
\begin{equation}
	\psi(x_t = s_i|x_{1...t-1}, z) \propto \exp((W_o g_t)_i)
\end{equation}
where $s_i$ is an element of the state space $S$, indexed by $i=1, ..., |S|$, and $W_o \in \mathbb{R}^{k \times |S|}$ is a parameter.

@@

\paragraph{Optimization}
We optimize the parameters of the neural networks expressing $\phi, \psi, q$ for (\ref{eq:bound}) using Backpropagation and Adam \citep{kingma-adam:-2014}, a standard and widely used method for optimizing neural networks.
During optimization, we approximate the gradient by taking a single sample from $z$ per sample trajectory and use the reparametrized gradient estimator introduced by \citet{kingma-auto-encoding-2014}.
%\begin{equation}\label{eq:bound-mc}
%	\partial_{\phi, \psi, q}	\log P(X_{0\dots T} | z_{-M...0}) + \beta \cdot \log \frac{N(\mu, \sigma^2)(z)}{q(z_{-M...0})}
%\end{equation}
%for a given training example $x_{-M...M}$.
This results in an unbiased estimator of the gradient of (\ref{eq:bound}) w.r.t. the parameters of $\phi, \psi, q$.

Following standard practice in machine learning, we split the data into a training and a held-out set, using the training set for optimizing the parameters.
After every pass through the training set, the objective is evaluated on the held-out set; optimization terminates once the value on the held-out set does not improve any more.

\paragraph{Evaluating}
After optimizing the parameters on a set of observed trajectories, we evaluate it on a held-out set of trajectories to estimate (\ref{eq:bound}).
Given parameters for $\phi, \psi, q$, we evaluate the objective~(\ref{eq:bound}) on the dataset by taking, for each timeseries $X_{-M}...X_M$, a single sample of $z  \sim \mathcal{N}(\mu, \sigma^2)$ and computing
\begin{equation}\label{eq:bound-mc}
	\frac{1}{N}	\sum_{X_{-M...M} \in TestData}	\log P(X_{0\dots M} | z_{-M...-1}) + \beta \cdot \log \frac{\mathcal{N}(\mu, \sigma^2)(z)}{q(z)}
\end{equation}
When evaluating (\ref{eq:bound}) on unseen data this way, the resulting value is guaranteed to be an \emph{upper} bound on the true predictive rate-distortion due to (\ref{ineq1}-\ref{ineq2}), up to sampling error introduced by sampling $z$ and the finite size of the held-out dataset.
It is important to note that this sampling error is different from the overfitting issue affecting OCE: 
(\ref{eq:bound-mc}) is an \emph{unbiased} estimator of an upper bound.
The magnitude of the estimation error can be estimated using standard methods such as bootstrapping, cross-validation, and taking more samples of $z$---unlike overfitting affecting OCE, which \emph{biases} the values obtained.
%The use of a held-out set is motivated by the fact that -- as for OCE -- the codebook may have overfitted to the set used for optimization.

@@

NPRD provably provides \emph{upper bounds} on the predictive rate-distortion curve.
This makes it possible to objectively compare the quality of different estimation methods:
Among methods that provide upper bounds, the one that provides the lowest such bound for a given problem is the one giving results closest to the true curve. \rljf{Does the sampling error issue arise here?}

@@

%We optimize $q(z|x)$, marginal $s(z)$, and $r(y|z)$.
\paragraph{OCE}
As discussed above, OCE overestimates structure in the process due to overfitting.
To address this, and to enable fair comparison between OCE and our neural method, we follow the evaluation method used in our neural method:
We use held-out data to compute \rljf{Describe these in prose too and what they mean}
$$\frac{1}{N} \sum_{i=1}^N D_{KL}(q(\cdot|x_i) || s(\cdot))$$ %\E_{z \sim q(\cdot|x_i)} \log \frac{q(z|x_i)}{s(z)} $$
and
$$\frac{1}{N} \sum_{i=1}^N \E_{z \sim q(\cdot|x_i)} \log r(y_i|z).$$
Thanks to (\ref{ineq1}-\ref{ineq2}), this provides upper bounds, up to sampling error introduced by finiteness of the held-out data.
Again, this kind of error does not bias the results in either direction, unlike overfitting, which introduces a systematic bias.

In order to be able to apply OCE on held-out data, which will contain sequences that did not occur during estimation, we add a pseudo-sequence $\omega$ and add pseudo-observations $(\omega, x), (x, \omega), (\omega, \omega)$ for all observed sequences $x_{-M...0}, x_{1...M}$ to the matrix of observed counts that serves as the input to the Blahut-Arimoto algorithm.
These pseudo-observations were assigned pseudo-counts $\gamma$ in this matrix of observed counts; we found that a wide range of values ranging from $0.0001$ to $1.0$ yielded essentially the same results.
When evaluating the codebook on held-out data, novel sequences were mapped to $\omega$.
%Given that all unseen futures were mapped to $\omega$, this method may slightly bias towards underestimating entropy of the

@@

Neural networks have hyperparameters, such as the number of units and the step size used in optimization, that affect the quality of approximation depending on properties of the dataset and the function being approximated.
Given that NPRD provably provides upper bounds, one can identify the best hyperparameters for a given problem by choosing the combination that leads to the lowest estimated upper bounds.

@@

\begin{tabular}{llll}
	            & Tractable\rljf{??} \\
	Dropout rate & 0.0, 0.1, 0.2 \\
	Embedding Dimension & 50, 100 \\
	LSTM Dimension & 32 \\
	LSTM Layers & 1 \\
	Learning Rate & 0.0001, 0.0005, 0.001 (4 times), 0.002, 0.004 \\
	Input Dropout & 0 \\
	Batch Size & 16, 32, 64 \\
	Flow Length & 1, 2 \\
	Flow Type & DSF, DDSF \\
	Flow Dimension & 32, 64, 128, 512 \\
	Flow Layers & 2 \\
\end{tabular}

@@

We first test NPRD on two processes where the predictive rate-distortion tradeoff is analytically tractable. % by computing the epsilon machine \citep{marzen-predictive-2016}. %, and where prior work has found the epsilon machine.
The \emph{Even Process} \citep{marzen-predictive-2016} is the process of 0/1 IID coin flips, conditioned on all blocks of consecutive ones having even length.
Its complexity and excess entropy are both $\approx 0.63$ nats. % (stationary distribution: 1/3 for odd, 2/3 for even)
It has infinite Markov order, and \cite{marzen-predictive-2016} find that OCE (at $M=5$) performs poorly. %, overestimating  even at $M=5$.
%Their method using the (already known) epsilon machine  works well.
The true predictive rate-distortion curve was computed in \cite{marzen-predictive-2016} using the analytically known epsilon machine.
The \emph{Random Insertion Process} \citep{marzen-predictive-2016} consists of sequences of ones with blocks of at most two zeros interspersed.
The $\epsilon$-machine is given in Figure~\ref{fig:rip-machine} (right).

@@

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/even-info.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/rip-info.pdf}

%\includegraphics[width=0.45\textwidth]{code/figures/even-beta-mem.pdf}
	\caption{Rate-Distortion for the Even Process and the Random Insertion Process. Gray lines: Analytical curves. Red dots: Multiple runs of NPRD. Red line: Tradeoff curve computed from NPRD runs. Blue: OCE for $M\leq 5$. }\label{fig:even}
\end{figure*}

@@

\rljf{Describe the RIP}

@@

In Figure~\ref{fig:rip-machine}, we record, for each of the three modes where the distribution of the code $Z$ shifts when a symbol is appended.
For comparison, we show the $\epsilon$-machine computed by \cite{marzen-predictive-2016}.
Comparing the two diagrams shows that NPRD effectively recovers the $\epsilon$-machine. \rljf{More detail on this part: it's a bit telegraphic}% even though $Z$ is a multivariate Gaussian variable.

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_2.png}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3.png}
	\caption{Applying PCA to 5000 sampled codes $Z$ for the Random Insertion Process, at $\beta = 0.25$ and $\beta = 0.6$. Samples are colored according to the states in the $\epsilon$ machine. Black samples belong to sequences that, at $M=15$, cannot be uniquely attributed to any of the states (ambiguous between A and C).}\label{fig:latent}
\end{figure*}
% colors = [{"A" : "red", "B" : "green", "C" : "blue", "U" : "black"}[x] for x in states]

\begin{figure*}
\includegraphics[width=0.45\textwidth]{code/figures/foo_pca_3_machine.png}
\begin{tikzpicture}[shorten >=1pt,node distance=4cm,on grid,auto] 
   \node[state, fill=green] (B)   {$B$}; 
   \node[state, fill=red] (A) [below left=of B] {$A$}; 
   \node[state, fill=blue] (C) [below right=of B] {$C$}; 
    \path[->] 
	(A) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (B)
	(A) edge [bend left=20] node [below] {$\frac{1}{2} | 1$} (C)
	(B) edge [bend left=20] node [above] {$\frac{1}{2} | 0$} (C)
	(B) edge [bend right=20] node [left] {$\frac{1}{2} | 1$} (C)
	(C) edge [bend left=20] node [left] {$\frac{1}{2} | 1$} (A)
	;
%
%%          edge  node [swap] {1} (B)
%    (C) edge  node  {1} (A)
%          edge [loop above] node {0} ()
%	(A) edge  node [swap] {0} (C) 
%          edge [loop below] node {1} ();
\end{tikzpicture}

@@

We consider the following process, which we'll call \textsc{Copy3}: $X_{-15}, ..., X_{-1}$ are independent uniform draws from $\{1,2,3\}$, and $X_1 = X_{-1}, ..., X_{15} = X_{-15}$.
This process deviates a bit from our usual setup since we defined it only for $t \in \{-15, ..., 15\}$, but it is well-suited to investigating this question:
The number of causal states is $3^{15} \approx 14$ million.
With exactly the same setup as for the \textsc{Even} and \textsc{RIP} processes, NPRD achieved essentially zero distortion on unseen data, even though the number of training samples (3 Million) was far lower than the number of distinct causal states.
However, we found that, at this setup, NPRD overestimated the rate.
Increasing the number of training samples from 3M to 6M, NPRD recovered codebooks that achieved both almost zero distortion and almost optimal rate, on fresh samples (Figure~\ref{fig:repeat}).
Even then, the number of distinct causal states is more than twice the number of training samples.
These results demonstrate that, by using function approximation, NPRD is capable of extrapolating to unseen causal states, encoding and decoding appropriate codes on the fly.

Note that one could easily design an optimal decoder and encoder for \textsc{Copy3} by hand -- the point of this experiment is to demonstrate that NPRD is capable of inducing such a codebook purely from data, in a general-purpose, off-the-shelf manner.
This contrasts with OCE:
Without optimizations specific to the task at hand, a direct application of OCE would require brute-force storing of all 14 million distinct pasts and futures.

@@

\begin{figure*}
%\includegraphics[width=0.45\textwidth]{code/figures/repeat2-ee-mem.pdf}
\includegraphics[width=0.45\textwidth]{code/figures/repeat3-ee-mem-frontier.pdf}
	\caption{Left: Rate-Distortion for the \textsc{Copy3} process. Green: Curve found by NPRD. Gray: Anaytical curve. \rljf{Red not green? Also y axis should be Memory, not Memories. Also let's label as Memory (nats) throughout (or bits), and Excess Entropy on the x axis throughout.}.}\label{fig:repeat}
\end{figure*}

@@

We consider the problem of estimating rate-distortion for Natural Language. \rljf{Some motivation for why this is cool and interesting, maybe in parallel with the Bentz et al entropy paper, and other information theoretic linguistic complexity papers like Koplenig et al, and Dembowski's work}
%
%As (CITE) have noted, evaluating (X) on finite data will overestimate MI with the future.
%To address this, we estimate (X) on held-out data.
%
%We compare our estimator with OCF on (1) POS-level language modeling, (2) word-level language modeling
%
%Challenges for OCF:
%
%- held-out data will have unobserved sequences.
%We add an out-of-vocabulary pseudo-sequence to the table of observed sequences.
%
